\chapter{Special-purpose components}
\index{Library!Components!misc}

The chapter deals with components that are not easily included
in any of the other chapters because of their special nature,
but which are still part of the \MCS system.

One part of these components deals with splitting simulations
into two (or more) stages. For example, a guide system is often
not changed much, and a long simulation of neutron rays
``surviving'' through the guide system could be reused
for several simulations of the instrument back-end, speeding up
the simulations by (typically) one or two orders of magnitude.
The components for doing this trick is {\bf Virtual\_input} and
{\bf Virtual\_output}, which stores and reads neutron rays, respectively.

Other components perform the simulation of the instrument
resolution functions. These are {\bf Res\_sample} and {\bf TOF\_Res\_sample},
which are to be
placed at the sample position, and {\bf Res\_monitor}, that should
be localized at the position of the instrument detector.

{\bf Progress\_bar} is a simulation utility that displays the simulation
status, but assumes the form of a component.

\newpage
\input{sources/Virtual_output}

\input{sources/Virtual_input}

\newpage
\input{misc/Res_sample}

\input{misc/TOFRes_sample}

\input{misc/Res_monitor}

\newpage
\section{Progress\_bar: Simulation progress and automatic saving}
\component{Progress\_bar}{System}{percent, flag\_save, profile}{}{}
\label{s:progress-bar}
\index{Simulation progress bar}

This component displays the simulation progress and status
but does not affect the neutron parameters.
The display is updated in regular intervals of the full simulation;
the default step size is 10 \%, but it may be changed using
the \verb+percent+ parameter (from 0 to 100).
The estimated computation time is displayed at the begining
and actual simulation time is shown at the end.

Additionally, setting the \verb+flag_save+ to 1 results in
a regular save of the data files during the simulation.
This means that is is possible to view the data before the end
of the computation, and have also a trace of it in case of
computer crash. The achieved percentage of the simulation is stored in these temporary
data files. Technically, this save is equivalent to sending regularly
a USR2 signal to the running simulation.

The optional 'profile' parameter, when set to a file name, will produce the number of statistical events reaching each component in the simulation. This may be used to identify positions where events are lost.

\section{Beam\_spy: A beam analyzer}
\component{Beam\_spy}{System}{}{}{should overlap previous component}
\index{Monitors!Beam analyzer}

This component is at the same time an Arm and a simple Monitor. It analyzes all neutrons reaching it, and computes statistics for the beam, as well as the intensity.

This component does not affect the neutron beam, and does not contain any propagation call. Thus it gets neutrons from the previous component in the instrument description, and should better be placed at the same position, with \verb+AT (0,0,0) RELATIVE PREVIOUS+.
