\section{Phonon\_simple: A simple phonon sample}
\label{s:phonon_simple}
\index{Samples!Phonon scattering}
\index{Inelastic scattering}

\component{Phonon\_simple}{Kim Lefmann, Ris\o\ National Laboratory}{ $r_{\rm o}$, $h$, $r_{\rm foc}$, $x_{\rm target}$, $y_{\rm target}$, $z_{\rm target}$, $\sigma_{\rm abs}$, $\sigma_{\rm inc}$, $a$, $b$, $c$, $M$, $DW$, $T$}{$w_x$, $h_y$, $t_z$, $w_{\rm focus}, h_{\rm focus}$, $w_{\rm foc, angle}$, $h_{\rm foc, angle}$, target\_index}{only validated qualitatively}

This component models a simple phonon signal from a single crystal of
a pure element in an {\em fcc} crystal structure.
Only one isotropic acoustic phonon branch is modelled, and the longitudinal
and transverse dispersions are identical with the velocity of sound being $c$.
Other physical parameters are the atomic mass, $M$, the lattice parameter, $a$,
the scattering length, $b$,
the Debye-Waller factor, \verb+DW+, and the temperature, $T$.
Incoherent scattering and absorption are taken into account by the cross
sections $\sigma_{\rm abs}$ and $\sigma_{\rm inc}$.

The sample can have the form of a cylinder with height $h$ and radius
$r_0$, or a box with dimensions $w_x, h_y, t_z$.

Phonons are emitted into a specific range of solid angles, specified
by the location $(x_t, y_t, z_t)$ and the focusing radius, $r_0$.
Alternatively, the focusing is given by a rectangle,
$w_{\rm focus}$ and $h_{\rm focus}$, and the focus point is given by the
index of a down-stream component, \verb+target_index+.

Multiple scattering is not included in this component.

A usage example of this component can be found in the \verb+Neutron site/tests/Test_Phonon+ instrument from the \verb+mcgui+.

\subsection{The phonon cross section} % This is modified from the paper version %
The inelastic phonon cross section for a Bravais crystal of a pure element
is given by Ref.~\cite[ch.3~]{squires}
\begin{eqnarray}
\frac{d^2\sigma'}{d\Omega dE_{\rm f}} &=&
  b^2 \frac{k_{\rm f}}{k_{\rm i}} \frac{(2\pi)^3}{V_0}\frac{1}{2M} \exp(-2W) \nonumber \\
&\times&
  \sum_{\tau,q,p} \frac{(\mbox{\boldmath $\kappa$} \cdot {\bf e}_{q,p})^2}
                       {\omega_{q,p}}
  \left\langle n_{q,p} + \frac{1}{2} \mp \frac{1}{2} \right\rangle
  \delta(\omega\pm\omega_{q,p}) \delta(\kappa\pm{\bf q}-\tau) ,
\end{eqnarray}
where both annihilation and creation of one phonon is considered
(represented by the plus and minus sign in the dispersion delta functions,
respectively).
In the equation,
$\exp(-2W)$ is the Debye-Waller factor, \verb+DW+ and
$V_0 $ is the volume of the unit cell.
The sum runs over the reciprocal lattice vectors, $\tau$,
over the polarisation index, $p$,
and the $N$ allowed wave vectors {\bf q} within the Brillouin zone
(where $N$ is the number of unit cells in the crystal).
Further, ${\bf e}_{q,p}$ is the
polarization unit vectors, $\omega_{q,p}$ the phonon dispersion,
and the Bose factor is
$\langle n_{q,p} \rangle = (\hbar \exp(|\omega_{q,p}|/k_{\rm B}T)-1)^{-1}$.

We have simplified this expression by assuming no polarization
dependence of the dispersion, giving
$\sum_{p} (\mbox{\boldmath $\kappa$} \cdot {\bf e}_{q,p})^2 = \kappa^2$.
We assume that the inter-atomic interaction is nearest-neighbour-only
so that the phonon dispersion becomes:
\begin{equation}
d_1({\bf q}) = c_1/a \sqrt{z-s_q} ,
\end{equation}
where $z=12$ is the number of nearest neighbours and
$s_q=\sum_{\rm nn} \cos({\bf q} \cdot {\bf r}_{\rm nn})$,
where in turn ${\bf r}_{\rm nn}$ is the lattice positions of the
nearest neighbours.

This dispersion relation may be modified with a small effort,
since it is given as a separate c-function attatched to the component.

To calculate $d\sigma/d\Omega$ we need to transform the
{\bf q} sum into an integral over the Brillouin zone by
$\sum_q \rightarrow N V_{\rm c} (2\pi)^{-3} \int_{\rm BZ} d^3{\bf q}$.
The $\mbox{\boldmath $\kappa$}$ sum can now be removed by
expanding the {\bf q} integral to infinity.
All in all, the partial differential cross section reads
\begin{eqnarray}
\frac{d^2\sigma'}{d\Omega dE_{\rm f}}
  (\mbox{\boldmath $\kappa$},\omega) &=&
  N b^2 \frac{k_{\rm f}}{k_{\rm i}} \frac{1}{2M}
  \int \frac{\hbar \kappa^2}{\hbar \omega_q}
  \left\langle n_{q}+\frac{1}{2}\mp\frac{1}{2} \right\rangle
  \delta(\omega\pm\omega_{q}) \delta(\mbox{\boldmath $\kappa$}\pm{\bf q})
   d^3{\bf q} \nonumber \\
 &=& N b^2 \frac{k_{\rm f}}{k_{\rm i}}
          \frac{\hbar^2 \kappa^2}{2M \hbar \omega_q}
  \left\langle n_{\kappa}+\frac12\pm\frac12 \right\rangle
  \delta(\hbar\omega\pm d_1(\kappa)) . \label{e:phonon-pdcross}
\end{eqnarray}

\subsection{The algorithm}
All neutrons, which hit the sample volume, are scattered
into a particular range of solid angle, $\Delta \Omega$,
like many other components. One of the difficult things in
scattering from a dispersion is to take care to fulfill the
dispersion criteria and to find the correct weight transformation.

In {\bf Phonon\_simple}, the following steps are taken:
\begin{enumerate}
\item If the sample is hit, calculate the total path length inside the
sample, otherwise leave the neutron ray unchanged.
\item Choose a scattering point inside the sample
\item Choose a direction for the final wave vector, $\hat{\bf k}_{\rm f}$
within $\Delta\Omega$.
\item Calculate possible values of $k_{\rm f}$ so that the
dispersion relation is fulfilled for the corresponding value
of ${\bf k}_{\rm f}$. (There is always at least one possible $k_{\rm f}$
value \cite{bacon}.)
\item Choose one of the calculated $k_{\rm f}$ values.
\item Propagate the neutron to the scattering point and adjust the
neutron velocity according to $k_{\rm f}$.
\item Calculate and apply the correct weight factor correction, see below.
\end{enumerate}

\subsection{The weight transformation}
Before making the weight transformation, we need to calculate the
probability for scattering along one certain direction $\Omega$
from one phonon mode. To do this, we must integrate out the delta
functions in the cross section (\ref{e:phonon-pdcross}).
We here use that $\hbar \omega_q = \hbar^2 (k_i^2 - k_f^2) / (2 m_{\rm N})$,
$\kappa = {\bf k}_{\rm i} - k_{\rm f}\hat{\bf k}_f$, and
the integration rule $\int \delta(f(x)) = (df/dx)(0)^{-1}$.
Now, we reach
\begin{equation} \label{eq:phononcross}
\left(\frac{d\sigma'}{d\Omega}\right)_j = \int \frac{d^2\sigma'}{d\Omega dE_{\rm f}} dE_{\rm f}
 = N b^2 \frac{k_{\rm f}}{k_{\rm i}}
\frac{\hbar^2 \kappa^2}{2M d_1(\kappa_j) J(k_{{\rm f},j})}
\left\langle n_{\kappa}+\frac12\pm\frac12 \right\rangle .
\end{equation}

where the Jacobian reads
\begin{equation}
J = 1 - \frac{m_{\rm N}}{k_{\rm f} \hbar^2}
    \frac{\partial}{\partial k_{\rm f}} \left( d_1(\kappa) \right) .
\end{equation}

A rough order-of-magnitude consideration gives
$\frac{k_{{\rm f},j}}{k_{\rm i}}\approx 1$,
$J \approx 1$,
$\langle n_{\kappa}+\frac12\pm\frac12 \rangle \approx 1$,
$\frac{\hbar^2\kappa^2}{2M d_1(\kappa)}
\approx \frac{m}{M}$.
Hence, $\left(\frac{d\sigma}{d\Omega}\right)_j \approx N b^2 \frac{m}{M}$, and
the phonon cross section becomes a fraction of
the total scattering cross section $4 \pi N b^2$, as it must be.
The differential cross section per unit volume is found from
(\ref{eq:phononcross}) by replacing $N$ with $1/V_0$.

The total weight transformation now becomes
\begin{equation} \label{eq:phonon_mult}
\pi_i = a_{\rm lin} l_{\rm max} n_{\rm s} \Delta \Omega
 b^2 \frac{k_{{\rm f},j}}{k_{\rm i}}
 \frac{\hbar^2 \kappa}{2 V_0 M d_1(\kappa) J(k_{{\rm f},j})}
 \left\langle n_{\kappa}+\frac12 \pm\frac12 \right\rangle ,
\end{equation}
where $n_s$ is the number of possible dispersion values in the chosen direction.

The \verb+Test_Phonon+ test/example instrument exists in the distribution for this component.
