% Emacs settings: -*-mode: latex; TeX-master: "manual.tex"; -*-

\def\textbfMCRC#1{\indexMCRC{#1}{textit}\textbf{#1}}
\def\textbfMCRH#1{\indexMCRH{#1}{textit}\textbf{#1}}
\def\textbfMCCC#1{\indexMCCC{#1}{textit}\textbf{#1}}
\def\textbfMCCH#1{\indexMCCH{#1}{textit}\textbf{#1}}
\def\textbfRTLC#1{\indexRTLC{#1}{textit}\textbf{#1}}

\chapter{Libraries and constants}
\label{c:kernelcalls}
\index{Library|textbf}

\indexLIB{mcstas-r}{textbf}
\indexLIB{mccode-r}{textbf}
\index{Preprocessor macros}
\index{C functions}

The \MCS Library contains a number of built-in functions
and conversion constants which are useful when constructing
components. These are stored in the \verb+share+ directory of
the \verb+MCSTAS+ library. \index{Library!components!share}
\indexEV{MCSTAS}

Within these functions, the 'Run-time' part is available for all
component/instrument descriptions. The other parts
% (see table~\ref{t:comp-share})
are dynamic, that is they are not
pre-loaded, but only imported once when a component requests it
using the \verb+%include+ \MCS keyword. For instance, within a
component C code block, (usually SHARE or DECLARE):
\indexKW{\%include}
\begin{lstlisting}
    %include "read_table-lib"
\end{lstlisting}
will include the 'read\_table-lib.h' file, and the 'read\_table-lib.c'
(unless the \verb+--no-runtime+ option is used with \verb+mcstas+).
Similarly,
\begin{lstlisting}
    %include "read_table-lib.h"
\end{lstlisting}
will \emph{only} include the 'read\_table-lib.h'.
The library embedding is done only once for all components (like the
 SHARE section). \indexKW{SHARE} For an example
of implementation, see \textbf{Res\_monitor}.

In this Appendix, we present a short list of both each of the library contents
and the run-time features.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Run-time calls and functions (\texttt{mcstas-r})}
\label{s:calls:run-time}
\index{Library!run-time|textbf}
Here we list a number of preprogrammed macros
which may ease the task of writing component and instrument definitions.

%-------------------------------------------------------------------------------
\subsection{Neutron propagation}

\def\textbfMCRH#1{\indexMCRH{#1}{textit}\textbf{#1}}
Propagation routines perform all necessary operations to transport neutron rays
from one point to an other. Except when using the special
\verb+ALLOW_BACKPROP;+ call prior to executing any \verb+PROP_*+ propagation,
the neutron rays which have negative propagation times are removed automatically.
\begin{itemize}
\item \textbfMCRH{ABSORB}. This macro issues an order to the overall
  \MCS simulator to interrupt the simulation of the current neutron
  history and to start a new one.
\item \textbfMCRH{PROP\_Z0}. Propagates the neutron to the $z=0$ plane,
  by adjusting $(x,y,z)$ and $t$ accordingly from knowledge of the
  neutron velocity $(vx,vy,vz)$.
  If the propagation time is negative, the neutron ray is absorbed, except if a \verb+ALLOW_BACKPROP;+ preceeds it.

  For components that are centered along the $z$-axis,
  use the \verb+_intersect+ functions to determine intersection time(s),
  and then a \verb+PROP_DT+ call.
\item \textbfMCRH{PROP\_DT}$(dt)$. Propagates the neutron through the
  time interval $dt$, adjusting $(x,y,z)$ and $t$ accordingly
  from knowledge of the neutron velocity. This macro automatically calls PROP\_GRAV\_DT when the \verb+--gravitation+ option has been set for the whole simulation.
\item \textbfMCRH{PROP\_GRAV\_DT}$(dt,Ax,Ay,Az)$. Like \textbf{PROP\_DT}, but it also
  includes gravity using the acceleration $(Ax,Ay,Az)$. In addition
  to adjusting $(x,y,z)$ and $t$, also $(vx,vy,vz)$ is modified.
\item \textbfMCRH{ALLOW\_BACKPROP}. Indicates that the next propagation routine
  will not remove the neutron ray, even if negative propagation times
  are found. Subsequent propagations are not affected.
  \index{Removed neutron events}
\item \textbfMCRH{SCATTER}. This macro is used to denote a scattering event
  inside a component.
%, see section~\ref{s:comp-trace}.
  It should be used e.g
  to indicate that a component has interacted with the neutron ray
  (e.g. scattered or detected).
  This does not affect the simulation (see, however, \textbf{Beamstop}),
  and it is mainly used by the
  \verb+MCDISPLAY+ section and the \verb+GROUP+ modifier
%(see~\ref{s:trace} and \ref{s:comp-mcdisplay}).
  See also the SCATTERED variable (below).
  \indexKW{GROUP}
  \indexKW{MCDISPLAY}
  \indexKW{EXTEND}
\end{itemize}

%-------------------------------------------------------------------------------
\subsection{Coordinate and component variable retrieval}
\index{Coordinate!retrieval functions}

\begin{itemize}
\item \textbfMCCH{MC\_GETPAR}$(comp, outpar)$. This may be used in e.g. the FINALLY section of an
  instrument definition to reference the output parameters of a
  component.
% See page~\pageref{mcgetpar} for details.
\item \textbfMCCH{NAME\_CURRENT\_COMP} gives the name of the current component as a string.
\item \textbfMCCH{POS\_A\_CURRENT\_COMP} gives the absolute position of the
  current component. A component of the vector is referred to as
  POS\_A\_CURRENT\_COMP.$i$ where $i$ is $x$, $y$ or $z$.
\item \textbfMCCH{ROT\_A\_CURRENT\_COMP} and
  \textbfMCCH{ROT\_R\_CURRENT\_COMP} give the orientation
  of the current component as rotation matrices
  (absolute orientation and the orientation relative to
  the previous component, respectively). A
  component of a rotation matrix is referred to as
  ROT\_A\_CURRENT\_COMP$[m][n]$, where $m$ and
  $n$ are 0, 1, or 2 standing for $x,y$ and $z$ coordinates respectively.
\item \textbfMCCH{POS\_A\_COMP}$(comp)$ gives the absolute position
  of the component with the name {\em comp}. Note that
  {\em comp} is not given as a string. A component of the
  vector is referred to as POS\_A\_COMP$(comp).i$
  where $i$ is $x$, $y$ or $z$.
\item \textbfMCCH{ROT\_A\_COMP}$(comp)$ and
  \textbfMCCH{ROT\_R\_COMP}$(comp)$ give the orientation of the
  component {\em comp} as rotation matrices (absolute
  orientation and the orientation relative to its
  previous component, respectively). Note that {\em comp}
  is not given as a string. A component of  a rotation
  matrice is referred to as
  ROT\_A\_COMP$(comp)[m][n]$, where $m$ and $n$ are
  0, 1, or 2.
\item \textbfMCCH{INDEX\_CURRENT\_COMP} is the number (index) of the
       current component  (starting from 1).
\item \textbfMCCH{POS\_A\_COMP\_INDEX}$(index)$ is the absolute position of
  component $index$. \\
  POS\_A\_COMP\_INDEX (INDEX\_CURRENT\_COMP) is the same as \\
  POS\_A\_CURRENT\_COMP. You may use \\
  POS\_A\_COMP\_INDEX  (INDEX\_CURRENT\_COMP+1) \\
  to make, for instance, your
  component access the position of the next component (this is usefull for
  automatic targeting).  A component of the vector is referred to as
  POS\_A\_COMP\_INDEX$(index).i$ where $i$ is $x$, $y$ or $z$.
\item \textbfMCCH{POS\_R\_COMP\_INDEX} works the same as above,
  but with relative coordinates.
\item \textbfMCRH{STORE\_NEUTRON}$(index, x, y, z, vx, vy, vz, t, sx, sy,
sz, p)$ stores the current neutron state in the trace-history table,
in local coordinate system. $index$ is usually INDEX\_CURRENT\_COMP.
This is automatically done when entering each component of an
instrument.
\item \textbfMCRH{RESTORE\_NEUTRON}$(index, x, y, z, vx, vy, vz, t, sx, sy,
sz, p)$ restores the neutron state to the one at the input of the
component $index$. To ignore a component effect, use
RESTORE\_NEUTRON (INDEX\_CURRENT\_COMP, \\
$x, y, z, vx, vy, vz, t,
sx, sy, sz, p$) at the end of its TRACE section, or in its EXTEND
section. These neutron states are in the local component coordinate
systems.
\item \textbfMCCH{SCATTERED} is a variable set to 0 when entering
  a component, which is incremented each time a SCATTER event occurs.
  \indexMCRH{SCATTER}{}
  This may be used in the \verb+EXTEND+ sections to determine whether
  the component interacted with the current neutron ray.
\item \textbf{extend\_list}($n$, \&\textit{arr}, \&\textit{len},
  \textit{elemsize}). Given an array \textit{arr} with \textit{len}
  elements each of size \textit{elemsize}, make sure that the array is
  big enough to hold at least $n$ elements, by extending \textit{arr}
  and \textit{len} if necessary. Typically used when reading a list of
  numbers from a data file when the length of the file is not known in advance.
\item \textbf{mcset\_ncount}$(n)$. Sets the number of neutron histories to simulate to $n$.
\item \textbf{mcget\_ncount}(). Returns the number of neutron histories to simulate (usually set by option \verb+-n+).
\item \textbf{mcget\_run\_num}(). Returns the number of neutron histories that have been simulated until now.
\end{itemize}

%-------------------------------------------------------------------------------
\subsection{Coordinate transformations}
\begin{itemize}
\item \textbfMCCC{coords\_set}$(x,y,z)$ returns a Coord structure (like POS\_A\_CURRENT\_COMP) with $x$, $y$ and $z$ members.
\item \textbfMCCC{coords\_get}$(P,$ \&$x$, \&$y$, \&$z)$ copies the $x$, $y$ and
$z$ members of the Coord structure $P$ into $x,y,z$ variables.
\item \textbfMCCC{coords\_add}$(a,b)$, \textbf{coords\_sub}$(a,b)$, \textbf{
coords\_neg}$(a)$ enable to  operate on coordinates, and return the
resulting Coord structure.
\item \textbfMCCC{rot\_set\_rotation}(\textit{Rotation t}, $\phi_x, \phi_y, \phi_z$)
  Get transformation matrix for rotation
  first $\phi_x$ around x axis, then $\phi_y$ around y,
  and last $\phi_z$ around z. $t$ should be a 'Rotation' ([3][3] 'double' matrix).
\item \textbfMCCC{rot\_mul}\textit{(Rotation t1, Rotation t2, Rotation t3)} performs $t3 = t1 . t2$.
\item \textbfMCCC{rot\_copy}\textit{(Rotation dest, Rotation src)} performs $dest = src$ for Rotation arrays.
\item \textbfMCCC{rot\_transpose}\textit{(Rotation src, Rotation dest)} performs $dest = src^t$.
\item \textbfMCCC{rot\_apply}\textit{(Rotation t, Coords a)} returns a Coord structure which is $t.a$
\end{itemize}

%-------------------------------------------------------------------------------
\subsection{Mathematical routines}
\begin{itemize}
\item \textbfMCCH{NORM}$(x,y,z)$. Normalizes the vector $(x,y,z)$ to have
  length 1 $^*$.
\item \textbfMCCC{scalar\_prod}$(a_x,a_y,a_z, b_x,b_y,b_z)$. Returns the scalar
  product of the two vectors $(a_x,a_y,a_z)$ and $(b_x,b_y,b_z)$.
\item \textbfMCCC{vec\_prod}($a_x$,$a_y$,$a_z$, $b_x$,$b_y$,$b_z$, $c_x$,$c_y$,$c_z$). Sets
  $(a_x,a_y,a_z)$ equal to the vector product $(b_x,b_y,b_z) \times
  (c_x,c_y,c_z)$  $^*$.
\item \textbfMCCH{rotate}($x$,$y$,$z$, $v_x$,$v_y$,$v_z$, $\varphi$, $a_x$,$a_y$,$a_z$). Set
  $(x,y,z)$ to the result of rotating the vector $(v_x,v_y,v_z)$
  the angle $\varphi$ (in radians) around the vector $(a_x,a_y,a_z)$  $^*$.
\item \textbfMCCC{normal\_vec}($n_x$, $n_y$, $n_z$, $x$, $y$, $z$).
  Computes a unit vector $(n_x, n_y, n_z)$ normal to the vector
  $(x,y,z)$.$^*$
\item \textbfMCCC{solve\_2nd\_order}(\&$t_1$, \&$t_2$ $A$,  $B$,  $C$).
  Solves the 2$^{nd}$ order equation $At^2 + Bt + C = 0$ and returns
  the solutions into pointers *$t_1$ and  *$t_2$. if $t_2$ is
  specified as \verb+NULL+, only the smallest positive solution is
  returned in $t_1$.
\end{itemize}
($^*$ The experienced c-programmer may be puzzled that these routines
can return information without the use of \emph{pass by reference},
the reason is that these calls are implemented as macros /
\verb+#define+ wrapped functions.)

%-------------------------------------------------------------------------------
\subsection{Output from detectors}
Details about using these functions are given in the \MCS User Manual.
\begin{itemize}
\item \textbfMCCH{DETECTOR\_OUT\_0D}$(...)$. Used to output the results from a
  single detector. The name of the detector is output together
  with the simulated intensity and estimated statistical error. The
  output is produced in a format that can be read by \MCS front-end
  programs.
%See section~\ref{s:comp-finally} ??? for details.
\item \textbfMCCH{DETECTOR\_OUT\_1D}$(...)$. Used to output the results from a
  one-dimensional detector. Integrated intensities error etc. is also
  reported as for DETECTOR\_OUT\_0D.
%See section~\ref{s:comp-finally} for details.
\item \textbfMCCH{DETECTOR\_OUT\_2D}$(...)$. Used to output the results from a
  two-dimentional detector. Integrated intensities error etc. is also
  reported as for DETECTOR\_OUT\_0D.
%See section~\ref{s:comp-finally} for details.
\item \textbfMCCH{DETECTOR\_OUT\_3D}$(...)$. Used to output
  the results from a three-dimentional detector. Arguments are the same as
  in DETECTOR\_OUT\_2D, but with an additional $z$ axis.
  Resulting data files are treated as 2D data, but the 3rd dimension is
  specified in the $type$ field. Integrated intensities error etc. is also
  reported as for DETECTOR\_OUT\_0D.
\item \textbf{mcinfo\_simulation}\textit{(FILE *f, mcformat,
  char *pre, char *name)} is used to append the simulation parameters into file $f$
  (see for instance \textbf{Res\_monitor}).
  Internal variable $mcformat$ should be used as specified.
  Please contact the authors for further information.
\end{itemize}

%-------------------------------------------------------------------------------
\subsection{Ray-geometry intersections}
\begin{itemize}
\item \textbfMCRC{inside\_rectangle}(\&$x$, \&$y$, $x$, $xw$, $yh$).
  Return 1 if $-xw/2 \leq x \leq xw/2$ AND $-yh/2 \leq y \leq yh/2$.
  Else return 0.
\item \textbfMCCC{box\_intersect}(\&$t_1$, \&$t_2$, $x$, $y$, $z$, $v_x$, $v_y$, $v_z$,
  $d_x$, $d_y$, $d_z$). Calculates the (0, 1, or 2) intersections between
  the neutron path and a box of dimensions $d_x$, $d_y$, and $d_z$,
  centered at the origin for a neutron with the parameters
  $(x,y,z,v_x,v_y,v_z)$. The times of intersection are returned
  in the variables $t_1$ and $t_2$, with $t_1 < t_2$. In the case
  of less than two intersections, $t_1$ (and possibly $t_2$) are set to
  zero. The function returns true if the neutron intersects the box,
  false otherwise.
\item \textbfMCCC{cylinder\_intersect}(\&$t_1$, \&$t_2$, $x$, $y$, $z$, $v_x$, $v_y$, $v_z$,
  $r$, $h$).  Similar to \textbf{box\_intersect}, but using a cylinder of height $h$ and radius $r$,
  centered at the origin.
\item \textbfMCCC{sphere\_intersect}(\&$t_1$, \&$t_2$, $x$, $y$, $z$, $v_x$, $v_y$, $v_z$,
  $r$). Similar to \textbf{box\_intersect}, but using a sphere
  of radius $r$.
\end{itemize}

%-------------------------------------------------------------------------------
\subsection{Random numbers}
\begin{itemize}
\item \textbfMCCC{rand01}(). Returns a random number distributed uniformly between 0 and 1.
\item \textbfMCCC{randnorm}(). Returns a random number from a normal
  distribution centered around 0 and with $\sigma=1$. The algorithm used to
  sample the normal distribution is explained in Ref.~\cite[ch.7]{num_rep}.
\item \textbfMCCC{randpm1}(). Returns a random number distributed uniformly between -1 and 1.
\item \textbfMCCC{randtriangle}(). Returns a random number from a triangular distribution between -1 and 1.
\item \textbfMCCC{randvec\_target\_circle}(\&$v_x$, \&$v_y$, \&$v_z$, \&$d\Omega$,
  aim$_x$, aim$_y$, aim$_z$, $r_f$). Generates a random vector $(v_x, v_y,
  v_z)$, of the same length as (aim$_x$, aim$_y$, aim$_z$), which is
  targeted at a \emph{disk} centered at (aim$_x$, aim$_y$, aim$_z$) with
  radius $r_f$ (in meters), and perpendicular to the \emph{aim} vector.. All directions
  that intersect the circle are chosen with equal probability. The solid
  angle of the circle as seen from the position of the neutron is returned
  in $d\Omega$. This routine was previously called \textbf{randvec\_target\_sphere}
  (which still works).
\item \textbfMCCC{randvec\_target\_rect\_angular}(\&$v_x$, \&$v_y$, \&$v_z$,
  \&$d\Omega$, aim$_x$, aim$_y$, aim$_z$,$h, w, Rot$) does the same as
  randvec\_target\_circle but targetting at a rectangle with angular dimensions
  $h$ and $w$ (in \textbf{radians}, not in degrees as other angles). The
  rotation matrix $Rot$ is the coordinate system orientation in the absolute
  frame, usually ROT\_A\_CURRENT\_COMP.
\item \textbfMCCC{randvec\_target\_rect}(\&$v_x$, \&$v_y$, \&$v_z$,
  \&$d\Omega$, aim$_x$, aim$_y$, aim$_z$,$height, width, Rot$) is the same as
  randvec\_target\_rect\_angular but $height$ and $width$ dimensions are given
  in meters. This function is useful to e.g. target at a guide entry window
  or analyzer blade.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reading a data file into a vector/matrix (Table input, \texttt{read\_table-lib})}
\label{s:read-table}
\indexLIB{read\_table-lib}{textbf}

The \verb+read_table-lib+ provides functionalities for reading text
  (and binary) data files. To use this library,
  add a \verb+%include "read_table-lib"+ in your component definition
  DECLARE or SHARE section. Tables are structures of type \verb+t_Table+
  (see \verb+read_table-lib.h+ file for details):
\begin{lstlisting}
/* t_Table structure (most important members) */
double *data;     /* Use Table_Index(Table, i j) to get element [i,j] */
long    rows;     /* number of rows */
long    columns;  /* number of columns */
char   *header;   /* the header with comments */
char   *filename; /* file name or title */
double  min_x;    /* minimum value of 1st column/vector */
double  max_x;    /* maximum value of 1st column/vector */
\end{lstlisting}

Available functions to read \emph{a single} vector/matrix are:
\begin{itemize}
\item \textbfRTLC{Table\_Init}(\&$Table$, $rows$, $columns$) returns an allocated
  Table structure. Use $rows=columns=0$ not to allocate memory and return an empty table.
  Calls to Table\_Init are \emph{optional}, since initialization is being
  performed by other functions already.
\item \textbfRTLC{Table\_Read}(\&$Table$, $filename$, $block$)
  reads numerical block number
  $block$ (0 to concatenate all) data from \emph{text} file $filename$ into $Table$,
  which is as well initialized in the process.
  The block number changes when the numerical data changes its size,
  or a comment is encoutered (lines starting
  by '\verb+# ; % /+'). If the data could not be read,
  then $Table.data$ is NULL and $Table.rows = 0$.
  You may then try to read it using Table\_Read\_Offset\_Binary.
  Return value is the number of elements read.
\item \textbfRTLC{Table\_Read\_Offset}(\&$Table$, $filename$, $block$, \&\textit{offset}, $n_{rows}$)
  does the same as Table\_Read except that it starts at offset \textit{offset}
  (0 means begining of file) and reads $n_{rows}$ lines (0 for all).
  The \textit{offset} is returned as the final offset reached after
  reading the $n_{rows}$ lines.
\item \textbfRTLC{Table\_Read\_Offset\_Binary}(\&$Table$, $filename$, $type$,
  $block$, \&\textit{offset}, $n_{rows}$, $n_{columns}$) does the same as
  Table\_Read\_Offset, but also specifies the $type$ of the file (may
  be "float" or "double"), the number $n_{rows}$ of rows to read, each
  of them having $n_{columns}$ elements. No text header should be present
  in the file.
\item \textbfRTLC{Table\_Rebin}(\&$Table$) rebins all $Table$ rows with increasing, evenly spaced first column (index 0), e.g. before using Table\_Value. Linear interpolation is performed for all other columns. The number of bins for the rebinned table is determined from the smallest first column step.
\item \textbfRTLC{Table\_Info}$(Table)$ print information about the table $Table$.
\item \textbfRTLC{Table\_Index}($Table, m, n$) reads the $Table[m][n]$ element.
\item \textbfRTLC{Table\_Value}($Table, x, n$) looks for the closest $x$
  value in the first column (index 0), and extracts in this row the
  $n$-th element (starting from 0). The first column is thus the 'x' axis for the data.
\item \textbfRTLC{Table\_Free}(\&$Table$) free allocated memory blocks.
\item \textbfRTLC{Table\_Value2d}($Table$, $X$, $Y$) Uses 2D linear interpolation on a Table, from (X,Y) coordinates and returns the corresponding value.
\end{itemize}

Available functions to read \emph{an array} of vectors/matrices in a \emph{text} file are:
\begin{itemize}
\item \textbfRTLC{Table\_Read\_Array}($File$, \&$n$) read and split $file$
into as many blocks as necessary and return a \verb+t_Table+ array.
Each block contains a single vector/matrix. This only works for text files.
The number of blocks is put into $n$.
\item \textbfRTLC{Table\_Free\_Array}(\&$Table$) free the $Table$ array.
\item \textbfRTLC{Table\_Info\_Array}(\&$Table$) display information about all data blocks.
\end{itemize}

The format of text files is free. Lines starting by '\verb+# ; % /+' characters are considered to be comments, and stored in $Table.header$. Data blocks are vectors and matrices. Block numbers are counted starting from 1, and changing when a comment is found, or the column number changes. For instance, the file 'MCSTAS/data/BeO.trm' (Transmission of a Berylium filter) looks like:
\begin{lstlisting}
  # BeO transmission, as measured on IN12
  # Thickness: 0.05 [m]
  # [ k(Angs-1) Transmission (0-1) ]
  # wavevector multiply
  1.0500  0.74441
  1.0750  0.76727
  1.1000  0.80680
  ...
\end{lstlisting}
Binary files should be of type "float" (i.e. REAL*32) and "double" (i.e. REAL*64),
and should \emph{not} contain text header lines. These files are platform
dependent (little or big endian).

The $filename$ is first searched into the current directory (and all user additional locations specified using the \verb+-I+ option, see the 'Running \MCS ' chapter in the User Manual), and if not found, in the \verb+data+ sub-directory of the \verb+MCSTAS+ library location. \index{Library!components!data}
\indexEV{MCSTAS} This way, you do not need to have local copies of the \MCS Library Data files (see table~\ref{t:comp-data}).

A usage example for this library part may be:
\begin{lstlisting}
  t_Table Table;       // declare a t_Table structure
  char file[]="BeO.trm";  // a file name
  double x,y;

  Table_Read(&Table, file, 1);  // initialize and read the first numerical block
  Table_Info(Table);            // display table informations
  ...
  x = Table_Index(Table, 2,5);  // read the 3rd row, 6th column element
                                // of the table. Indexes start at zero in C.
  y = Table_Value(Table, 1.45,1);  // look for value 1.45 in 1st column (x axis)
                                // and extract 2nd column value of that row
  Table_Free(&Table);           // free allocated memory for table
\end{lstlisting}
Additionally, if the block number (3rd) argument of  \textbf{Table\_Read} is 0, all blocks will be concatenated.
The \textbf{Table\_Value} function assumes that the 'x' axis is the first column (index 0).
Other functions are used the same way with a few additional parameters, e.g. specifying an offset for reading files, or reading binary data.

This other example for text files shows how to read many data blocks:
\begin{lstlisting}
  t_Table *Table;       // declare a t_Table structure array
  long     n;
  double y;

  Table = Table_Read_Array("file.dat", &n); // initialize and read the all numerical block
  n = Table_Info_Array(Table);     // display informations for all blocks (also returns n)

  y = Table_Index(Table[0], 2,5);  // read in 1st block the 3rd row, 6th column element
                                   // ONLY use Table[i] with i < n !
  Table_Free_Array(Table);         // free allocated memory for Table
\end{lstlisting}

You may look into, for instance, the source files for
\textbf{Monochromator\_curved} or \textbf{Virtual\_input}
for other implementation examples.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Monitor\_nD Library}
\indexLIB{monitor\_nd-lib}{textbf}

This library gathers a few functions used by a set of monitors e.g. Monitor\_nD, Res\_monitor, Virtual\_output, etc.
It may monitor any kind of data, create the data files, and may display many geometries (for \verb+mcdisplay+).
Refer to these components for implementation examples, and ask the authors for more details.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Adaptive importance sampling Library}
\indexLIB{adapt\_tree-lib}{textbf}

This library is currently only used by the components \textbf{Source\_adapt}
and \textbf{Adapt\_check}. It performs adaptive importance sampling of neutrons for simulation efficiency optimization.
Refer to these components for implementation examples, and ask the authors for more details.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Vitess import/export Library}
\indexLIB{vitess-lib}{textbf}

This library is used by the components
\textbf{Vitess\_input} and \textbf{Vitess\_output},
as well as the \verb+mcstas2vitess+ utility.
\indexMCTOOL{mcstas2vitess}{}
Refer to these components for implementation examples, and ask the authors for more details.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Constants for unit conversion etc.}
\index{Units!constants and conversions|textbf}
\index{Conversion!of units}
\index{Constants|textbf}

The following predefined constants are useful for conversion
between units
\def\textvb#1{\textbf{#1}\index{#1 (constant)@\texttt{#1} (constant)|textbf}}
\begin{center}
\begin{tabular}{|l|c|p{0.29\textwidth}|p{0.252\textwidth}|}
\hline
Name & Value & Conversion from & Conversion to \\ \hline
\textvb{DEG2RAD} & $2 \pi / 360$ & Degrees & Radians \\
\textvb{RAD2DEG} & $360 / (2 \pi)$ & Radians & Degrees \\
\textvb{MIN2RAD} & $2 \pi / (360 \cdot 60)$
  & Minutes of arc & Radians \\
\textvb{RAD2MIN} & $(360\cdot 60) / (2 \pi)$
  & Radians & Minutes of arc \\
\textvb{V2K} & $10^{10} \cdot m_\mathrm{N}/\hbar$
  & Velocity (m/s) & \textbf{k}-vector (\AA$^{-1}$) \\
\textvb{K2V} & $10^{-10} \cdot \hbar / m_\mathrm{N}$
  & \textbf{k}-vector (\AA$^{-1}$) & Velocity (m/s) \\
\textvb{VS2E} & $m_\mathrm{N} / (2 e)$
  & Velocity squared (m$^2$ s$^{-2}$) & Neutron energy (meV) \\
\textvb{SE2V} & $\sqrt{2 e/m_\mathrm{N}}$
  & Square root of neutron energy (meV$^{1/2}$) & Velocity (m/s) \\
\textvb{FWHM2RMS} & $1/\sqrt{8\log(2)}$
  & Full width half maximum & Root mean square (standard deviation) \\
\textvb{RMS2FWHM} & $\sqrt{8\log(2)}$
  & Root mean square (standard deviation) & Full width half maximum \\
\textvb{MNEUTRON} & $1.67492 \cdot 10^{-27}$~kg
  & Neutron mass, $m_\mathrm{n}$ & \\
\textvb{HBAR} & $1.05459 \cdot 10^{-34}$~Js
  & Planck constant, $\hbar$ & \\
\textvb{PI} & $3.14159265...$
  & $\pi$ & \\
\textvb{FLT\_MAX} & 3.40282347E+38F
  & a big float value & \\
\hline
\end{tabular}
\end{center}
