% Emacs settings: -*-mode: latex; TeX-master: "manual.tex"; -*-

\addcontentsline{toc}{chapter}{\protect\numberline{}{Preface and acknowledgments}}

\chapter*{Preface and acknowledgments}
This document contains information on the Monte Carlo neutron
ray-tracing program \MCS version \version, building on the initial
release in October 1998 of version 1.0 as presented in
Ref. ~\cite{nn_10_20} and further developed though version 2.0 as
presented in Ref. ~\cite{mcs_ppf}.
The reader of this document is supposed to have some knowledge of neutron scattering,
whereas only little knowledge about simulation techniques is
required. In a few places, we also assume familiarity with the
use of the C programming language and UNIX/Linux.

If you don't want to read this manual in full, go directly to the brief introduction in chapter \ref{s:brief}.

It is a pleasure to thank Prof.~Kurt N.~Clausen, PSI, for his continuous
support to this project and for having initiated \MCS in the first
place. Essential support has also been given by Prof.~Robert McGreevy, ISIS.
We have also benefited from discussions with many other people in the neutron scattering
community, too numerous to mention here.

In case of errors, questions, or suggestions, do not hesitate to
contact the authors at \verb+mcstas-support@mcstas.org+
or consult the \MCS home page~\cite{mcstas_webpage}.
A special bug/request reporting service is available \cite{github_issue_webpage}.

\index{McStas!mailing lists}
\index{Mailing lists!mcstas-users}
\index{McStas!citing|textbf}
\index{Citing McStas|textbf}
If you \textbf{appreciate} this software, please subscribe to the \verb+mcstas-users@mcstas.org+ email list, send us a smiley message, and contribute to the package. We also encourage you to refer to this software when publishing results, with the following citations:
\begin{itemize}
\item{P. Willendrup, E. Farhi, E. Knudsen, U. Filges and K. Lefmann,
    Journal of Neutron Research, \textbf{17} (2014) 35.}
\item{K. Lefmann and K. Nielsen, Neutron News \textbf{10/3}, 20, (1999).}
\item{P. Willendrup, E. Farhi and K. Lefmann, Physica B, \textbf{350} (2004) 735.}
\end{itemize}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{\MCS \version\ contributors}
\index{McStas!contributors|textbf}
\index{Contributors|textbf}

Several people outside the core developer team have been contributing
to \MCS \version:
\begin{itemize}
\item Thanks to Jonas Stein from Uni Cologne for helping us modernize the \TeX\ documentation
\item Thanks to Esben Klinkby and Troels Sch\"onfeldt from DTU Nutech for helping
  Peter W  with modernizing the ESS\_moderator.comp
\item Thanks to R. Heenan for contributing the ISIS ISIS\_SANS2d instrument 
\item Thanks to Klaus Habicht and Markos Skoulatos, HZB for
  contributing the HZB\_FLEX instrument
\item Thanks to Morten Sales, HZB  for contributing the SEMSANS\_instrument
\item Thanks to Henrich Frielinghaus, FZJ for contributing the
  SANS\_benchmark2.comp component and the related FZJ\_BenchmarkSfin2
  test instrument
\item Thanks to Esko Oksanen, ESS for contributing .lau reflection
  lists for the macromolecular structures Rubedoxin and Perdeuterated pyrophosphatase
\end{itemize}
Thank you guys! This is what \MCS is all about!

Third party software included (only those distributed with) \MCS are:
\begin{itemize}
\item Strawberry Perl (Windows system only)
\item perl Math::Amoeba from John A.R. Williams \verb+J.A.R.Williams@aston.ac.uk+.
\item perl Tk::Codetext from Hans Jeuken \verb+haje@toneel.demon.nl+.
\item PGPLOT from Tim Pearson \verb+tjp@astro.caltech.edu+ (Windows
  and Mac OS X systems only).
\item perl-PGPLOT from Karl Glazebrook \verb+karl@astro.swin.edu.au+ (Windows
  and Mac OS X systems only).
\item PDL (Perl Data Language) from \verb+http://pdl.perl.org+ (Windows
  and Mac OS X systems only).
\item NXSLib from Mirko Boin \verb+boin@helmholtz-berlin.de+
\end{itemize}
%and special thanks to the following \MCS contributors:
%\begin{itemize}
%\item Christophe Taton from ENSIMAG/Grenoble for the MPI support
%\item Reynald Arnerin from Universite J. Fourrier, Grenoble for the complex geometry support in samples
%\end{itemize} 

The \MCS project has been supported by the European Union
through``XENNI / Cool Neutrons'' (FP4), ``SCANS'' (FP5),
``nmi3/MCNSI'' (FP6), ```nmi3-ii/E-learning'' and ``nmi3-ii/MCNSI7''
(FP7) ~\cite{nmi3_webpage,mcnsi_webpage}.
\MCS was supported directly from the construction project for the ISIS second
target station (TS2/EU), see~\cite{ts2_webpage}. Currently \MCS is 
supported through the Danish involvement in the \emph{Data Management
  and Software Center}, a subdivision of the 
European Spallation Source (ESS), see~\cite{ess_webpage} and the
European Union through ``SINE2020/WP3 e-elarning''
and and ``SINE2020/WP8 e-Tools'' (Horizon2020).
the home pages~\cite{sine2020_webpage}.

